\texttt{timeline}: A Time Series Visualization Platform

### 2016-06-15 docs/report/paper.tex

\begin{abstract}
TODO
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
%Motivation and goals (which problem you are solving for the chosen data)

@@

\section{Solution}\label{sec:solution}
%Description of your solution: methodology, algorithms, design, use case

@@

Another possibility for the data import in the standalone application use case would be that the data can be imported using a dialog in the web-interface.
However, since data sets can be quite large (\ie multiple thousand samples), the upload time could be substantive. 
Moreover, since it was not planned to support persistent storage of the data sets in a rational or document-based database (at least not for the first release), the usability of the platform would suffer due to repeated re-uploads of the data sets.

@@

 \item He creates one or more data set objects that corresponds to the problem he or she is working on and registers it within the library.

@@

The usage of the platform as library also allows a simpler integration of live data sources by simply wrapping the data steam into a class that implements all required live data set methods.
Altogether we think that this design choice supports a more natural and quicker workflow with the platform.

@@

\section{Conclusion and Outlook}\label{sec:conclusion}
% Discussion and outlook: what worked well, what could be improved

% one problem: resultion of time only in seconds due to unix time stamp

>>>>

\begin{abstract}
In this paper a web-based visualization platform for time series data is proposed. 
It is not only able to represent static temporal data and it properties but also data from live sources, such as sensors. 
Furthermore, it allows the visualization of the output of time series prediction methods.
The main aspects of this paper are the overall design, the use cases, and the outlook for possible enhancements of the application.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}

@@

\section{Solution}\label{sec:solution}

@@

Another possibility for the data import in the standalone application use case would be the possibility to import the data using a dialog in the web-interface.
However, since data sets can be quite large (\ie multiple thousand samples), the upload time could be substantive. 
Moreover, since it was not planned to support persistent storage of the data sets in a rational or document-based database (at least not for the first release), the usability of the platform would suffer due to repeated re-uploads of the data sets.

@@

 \item He or she creates one or more \texttt{timeline} data set objects that corresponds to the problem he or she is working on and registers it within the library.

@@

The usage of the platform as library also allows a simpler integration of live data sources by simply wrapping the data steam into a class that implements all required live data set methods and the possibility to simply include data sets from other web resources (\eg from a website that publishes stock market data).
Altogether we think that this design choice supports a more natural and quicker workflow with the platform.

@@

The visualization primitives that can be generated by the platform are the following:

\begin{enumerate}
 \item A simple time series plot, which plots the observations (i.e. the data points) against the time of the observations.
 It is possible to filter the time series in a certain time span and include a rolling mean and rolling standard deviation into this plot.
 This can be used to determine whether or not the time series is stationary \cite{Nason2006}.

 \item A extended version of the simple time series plot that allows the plotting of multiple time series plots in one figure to compare them directly.

 \item A extended version of the simple time series plot that works in the conjunction with live data sets such that it can be updated dynamically.

 \item A autocorrelation plot \cite{Box1976}, also known as autocorrelation function (ACF) plot, which are a statistical tool to proof seasonal effects in the data.

 \item A forecast evaluation plot, that can be used to inspect the performance a time series prediction method. 
 It is a extended version of the simple time series plot as well, which shows the multiple parts of the forecast data set in one plot. 
\end{enumerate}

The back-end was implemented in Python using Flask \cite{Flask} as web-framework to render the user interface, deliver data using the API, and provide a small web-server such that the application can be executed.
To store the time series data in the back-end the data analysis library Pandas \cite{McKinney2010} was used. 
The application user interface uses the responsive Bootstrap \cite{Bootstrap} front-end framework and the JavaScript visualization library MetricsGraphics.js \cite{MetricsGraphics} to render the visualization primitives.

\section{Conclusion and Outlook}\label{sec:conclusion}

Like in most other projects, there is a lot room for potential improvements and possible feature work.
One constraint, due to the implementation of the framework, is that it only can handle dates and time stamps in a resolution of seconds.
This is caused by the fact that all date and time related parameters are encoded using the UNIX time stamp (\ie the number of seconds since midnight January 1, 1970 UTC) before they are sent from the front-end to the back-end and vice versa.
This is required since dates are often part of URLs, for example, in API requests.
An implication of this is, of course, that it is only possible to handle time series data within a resolution of one second (\eg the period between two samples must be at least one second) in the \texttt{timeline} platform.
This could be problematic, especially for real-time applications, where events usually happen in the range of milliseconds.
However, there are potential solutions for this issue.
For example, one option would be to encode dates as milliseconds since January 1, 1970 or a more general, and probable better, solution would be the usage of a well-defined date time format that allows the reliable parsing of dates from strings, which could be transmitted instead of the UINIX time stamp integers.

Another possibility for feature work would be the creation of a Python package that can be uploaded to PyPI \cite{PyPI}, the Python package index.
This would require to write a installer using the Python \texttt{setuptools} module. 
Furthermore, the number of available visualizations is quite small at the time, and the existing plots could be more interactive (\eg brushing).
There are plenty of other plots that could be integrated into the platform, for example, seasonality plots.
However, not only additional visualizations could be added to the platform, but also algorithms and statistical tests that allow further examination of the data sets (\eg implementation of the Dickey--Fuller test \cite{Dickey1979}).

In conclusion we think that the \texttt{timeline} visualization platform could be a valuable tool when doing time series analysis.
It could easily integrated in existing workflows and is modular structured to allow enhancements. 
There is room for potential improvements, but the first version of this application lays a solid foundation for future use cases.
